The core transcriptome of mammalian placentas and the divergence of expression with placental shape

### 2016-11-30 manuscript/manuscript.Rnw
clarify that the metatherians are excluded

To address our questions, we 
sequenced and quantified the total RNA of mammalian placentas at term,
including \human, \bonobo (the sister
species to \textit{Pan troglodytes}), \smonkey,
\mouse, \galili, \carmeli, \dog, \cow, \elephant,
\armadillo, and \opossum.
These species represent four major branches of eutherian mammals
(Euarchontoglires, Laurasiatheria, Afrotheria, and Xenartha) as well
as marsupials. These placentas also represent a wide array of
placental morphologies as well as traits such as number and size of
offspring and length of gestation (\cref{tab:gestational_periods}). We
use this data, along with existing RNA-Seq data from \mouse
\cite{Wang.ea2011:surveyfornovelimprinted}, \human
\cite{Hughes.ea2015:Evaluatingintra-andinter-individual},
\horse
\cite{Coleman.ea2010:Structuralannotationofequine}, \sheep
\cite{Archibald.ea2010:sheepgenomereferencesequence}, and
\pig \cite{Ahanda.ea2012:Predictionofaltered3-} to: 1)
identify the non-housekeeping genes which are expressed in the
placentas of all species analyzed and therefore represent the
\emph{core functionality of the placenta}; 2) identify genes whose
expression level changes significantly in each of the morphology
types.

@@

The
\Sexpr{core.placenta.transcriptome.long[housekeeping==FALSE,length(unique(human_name))]}
genes comprising the core placenta transcriptome are shown in
\cref{fig:core_placenta_transcriptome} ordered by their median
transcription level in \ac{FPKM}. We considered a gene to be core to
the placenta transcriptome if a 1:1 ortholog of that gene was
expressed with $\text{FPKM} \ge 10$ in all species studied and was not
a human housekeeping gene according to
\citet{Eisenberg.ea2013:Humanhousekeepinggenesrevisited}. Genes which
meet this criteria but are housekeeping genes are depicted in
Supplemental \cref{fig:core_placenta_transcriptome_housekeeping}.
\Cref{fig:core_placenta_transcriptome_shape} shows the number of genes
which would be included in the core placenta set if the \ac{FPKM}
cutoff was set differently; for example, choosing an \ac{FPKM} cutoff
of 1 yields %
\Sexpr{cpt.distribution.shape[`Expression (FPKM)`>=1 & `Min. Exp. in`
  == '14 of 14 Species',max(num_genes)]} %
genes in the core transcriptome. %
In contrast to the shape of the number of genes included in the core
placenta transcriptome, the set of \human housekeeping genes expressed
all placentas is flatter (Supplemental
\cref{fig:core_housekeeping_transcriptome_shape}), and includes more
of the total fraction of genes. However, there is still a large
difference between the set of housekeeping genes which are expressed
in \human and those which are expressed in all placentas.

>>>>

To address our questions, we 
sequenced and quantified the total RNA of mammalian placentas at term,
including \human, \bonobo (the sister
species to \textit{Pan troglodytes}), \smonkey,
\mouse, \galili, \carmeli, \dog, \cow, \elephant,
\armadillo, and \opossum.
These species represent four major branches of eutherian mammals
(Euarchontoglires, Laurasiatheria, Afrotheria, and Xenartha) as well
as marsupials. These placentas also represent a wide array of
placental morphologies as well as traits such as number and size of
offspring and length of gestation (\cref{tab:gestational_periods}). We
use this data, along with existing RNA-Seq data from \mouse
\cite{Wang.ea2011:surveyfornovelimprinted}, \human
\cite{Hughes.ea2015:Evaluatingintra-andinter-individual},
\horse
\cite{Coleman.ea2010:Structuralannotationofequine}, \sheep
\cite{Archibald.ea2010:sheepgenomereferencesequence}, and
\pig \cite{Ahanda.ea2012:Predictionofaltered3-} to: 1)
identify the non-housekeeping genes which are expressed in the
placentas of all eutherian species analyzed and therefore represent the
\emph{core functionality of the placenta}; 2) identify genes whose
expression level changes significantly in each of the morphology
types.

@@

The
\Sexpr{core.placenta.transcriptome.long[housekeeping==FALSE,length(unique(human_name))]}
genes comprising the core placenta transcriptome are shown in
\cref{fig:core_placenta_transcriptome} ordered by their median
transcription level in \ac{FPKM}. We considered a gene to be core to
the placenta transcriptome if a 1:1 ortholog of that gene was
expressed with $\text{FPKM} \ge 10$ in all eutherian species studied
(excluding the metatherian, \opossum) the and was not a human housekeeping
gene according to
\citet{Eisenberg.ea2013:Humanhousekeepinggenesrevisited}. Genes which
meet this criteria but are housekeeping genes are depicted in
Supplemental \cref{fig:core_placenta_transcriptome_housekeeping}.
\Cref{fig:core_placenta_transcriptome_shape} shows the number of genes
which would be included in the core placenta set if the \ac{FPKM}
cutoff was set differently; for example, choosing an \ac{FPKM} cutoff
of 1 yields %
\Sexpr{cpt.distribution.shape[`Expression (FPKM)`>=1 & `Min. Exp. in`
  == '14 of 14 Species',max(num_genes)]} %
genes in the core transcriptome. %
In contrast to the shape of the number of genes included in the core
placenta transcriptome, the set of \human housekeeping genes expressed
all placentas is flatter (Supplemental
\cref{fig:core_housekeeping_transcriptome_shape}), and includes more
of the total fraction of genes. However, there is still a large
difference between the set of housekeeping genes which are expressed
in \human and those which are expressed in all placentas.
